---
title: "ARIMA MODEL"
author: "Elbrus Gasimov, Alberto Delgado Lopez"
date: "6/2/2021"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```


## Literature review 

During the analysis and after the analysis to check every step and final result below sources have been used

- https://datascienceplus.com/time-series-analysis-using-arima-model-in-r/
- https://otexts.com/fpp2/arima-r.html
- https://www.analyticsvidhya.com/blog/2015/12/complete-tutorial-time-series-modeling/
- https://blogs.oracle.com/ai-and-datascience/post/introduction-to-forecasting-with-arima-in-r
- https://rpubs.com/riazakhan94/arima_with_example


## Introduction 

The idea of this work is to study and analyse, using time series metrics, the most recent data on Covid-19 deaths in the state of California. The main goal here is to predict death cases for short term and long term. Will build ARIMA model and then compare all possible valid models and choose the best one to handle our forecast!

Additionally, while building the models, two main hypothesis will be solved:
- First main hypothesis: number of deaths will significantly decrease during next month - Using ARIMA
- Secondary hypothesis: number of deaths is highly related with the cases of a demographic group - Using ARDL

### Data info

Our Dataset is about Covid Cases in US, California. It is a daily data and it comes from the 'California health and human services open data portal':

https://data.chhs.ca.gov/dataset/covid-19-time-series-metrics-by-county-and-state

The data is divided in two files:

- "covid19cases_test.csv" with reported death and cases of Covid-19 in the state of California since 2020 until present (last update May 22, 2021), it also includes information about the area, population, total number of tests, positive tests... From here, we are only interested in the total number of deaths, not divided by area, but from the sum of all the areas (whole California).

- "covid19casesdemographics.csv" with reported number of death and cases of Covid-19 in the state of California since 2020 until present (last update May 22, 2021). This dataset is divided in demographic groups. We are intereseted in the daily data of Covid cases for each demographic group.

As we focus on ARIMA model to predict death cases for this file we will focuc on the first dataset, where will observe deaths for a year and then will predict for long ans short time period.

```{r}
#Libraries 
library(lmtest)
library(fBasics)
library(urca)
library(xts)
library(forecast)
library(quantmod)
# load needed packages
library(xts)
library(fBasics)
library(urca)
source("function_testdf2.R")
source("function_import_data_into_xts.R")
# libraries
requiredPackages = c("lmtest","fBasics","urca","xts","forecast", "quantmod", "reshape2")

for(i in requiredPackages){if(!require(i,character.only = TRUE)) install.packages(i)}
for(i in requiredPackages){if(!require(i,character.only = TRUE)) library(i,character.only = TRUE)}

```

### Lodading the Dataset.

Importing the dataset and exploring it a bitLets load the datasets, analyse the data, do a little bit of cleaning and reorganize variables.

```{r}
covid <- read.csv("main_case.csv", # name of the file
                header = T,  	# if first row doesn't contain variable names: header=F'
                sep = ",", 	  # sign used as columns separator
                dec = ".")	  # sign used as decimal place separator
head(covid)
tail(covid)

summary(covid)
str(covid)

covid$date <- as.Date(covid$date, "%Y-%m-%d")

covid<-covid[,c(1,6)]
covid<-na.omit(covid)
covid<-aggregate(covid[,-1], by = list(covid$date), FUN = sum)
names(covid) <- c("date","deaths")

head(covid)
tail(covid)
summary(covid)
str(covid)
```

## Times Series object

Let's create the xts objects


```{r}
head(covid)
covid <- xts(covid[,-1], order.by=covid$date)
names(covid) <- c("deaths")
plot(covid)

plot(as.zoo(covid))

diffcovid <- diff.xts(covid)
plot(diffcovid)

diffcovid$deaths <- tsclean(diffcovid$deaths)
plot(diffcovid)


df.test <- ur.df(diffcovid$deaths,
                 type = c('none'),
                 lags = 0)

summary((df.test))

resids_ <- df.test@testreg$residuals
bgtest(resids_~1, order = 1)
bgtest(resids_~1, order = 2)
bgtest(resids_~1, order = 3)
bgtest(resids_~1, order = 4)
bgtest(resids_~1, order = 5)
```

## Residuals check

To be sure that conlusions may be drawn we need to check whether there is no autocorrelation in residuals
adf.test performance

```{r}
adf.test<- ur.df(diffcovid$deaths,
                 type = c('none'),
                 lags = 5)
summary(adf.test)

resids_ <- adf.test@testreg$residuals
bgtest(resids_~1, order = 1)
bgtest(resids_~1, order = 2)
bgtest(resids_~1, order = 3)
bgtest(resids_~1, order = 4)
bgtest(resids_~1, order = 5)


testdf2(variable = diffcovid$deaths,
        test.type = 'nc',
        max.augmentations = 5,
        max.order = 5)
```

As we reject the H0 we comfirm that your variable is Stationary! 

## Alternative tests

To check the residuals there here are some test we can conduct! 

```{r}
pp.test.d <-ur.pp(diffcovid$deaths,
                  type =  c("Z-tau"),
                  model = c("constant"))

summary(pp.test.d)
```

pp test here comfirms that our variable is stationary. Generaly, the result of df and pp tests should be the same just like ours to be in the safe side howver if the results are different then we contine testing with the 3rd test
KPSS test has a different H0 which is ts is stationary


```{r}
kpps.test <- ur.kpss(diffcovid$deaths,
                     type = c('mu'))
summary(kpps.test)
```

As a result we get that we fail to reject H0 which is our variable is stationary our 5 percent rejecttion intervals starts from 0.463 till plus infinite
our t statistics value is 0.3022 means it is out of the rejection zone!!

The Box-Jenkins procedure
Initial identification of parameters p and q:we need to see if the residuals are autocorrelated or not! if not then
we have a good model we can use the model to calculate the forecast, however if residuals are autocorrelated then we need to increase p and q values
and estimate the  model again

## ACF and PACF stationary test

lets see ACF and PACF non- stationary variable 

### ACF
```{r}
acf(diffcovid$deaths, 
    lwd = 5, # line width
    col = "dark green", # color
    main = "ACF for Death")
```

### PACF

```{r}
pacf(diffcovid$deaths, 
     lwd = 5, # line width
     col = "dark green", # color
     main = "PACF for Death")
```

As we have daily data,then my approuch we be to see if we have any pattern repeating itsel
as i can see somehow there is a weekly pattern, which makes feel choosing p = 5 and q = 5
We have done all steps and it is time to check our model for this we will start with creating
the model and we will use Arima() ## as it is better than arima(), because it allows us to include constant term to the varibale


```{r}
arima1 <- Arima(covid$deaths,
                order = c(5,1,5),
                include.constant = TRUE)

summary(arima1)
```

This output is not exaclty what we want so lets check coeftest() function, from the lmtest package to test the significance of model parameters.

```{r}
coeftest(arima1)
```

From the results we can see that 5th lag from AR and 1st lag from MA is statisticly significiant now lets check if the model is valid!

Are residuals of arima111 model white noise? 
resid() function applied to model results returns residuals

```{r}
par(mfrow = c(2,1))  
acf(resid(arima1), 
    lag.max = 36,
    ylim = c(-0.1,0.1), 
    lwd = 5, col = "dark green",
    na.action = na.pass)
pacf(resid(arima1), 
     lag.max = 36, 
     lwd = 5, col = "dark green",
     na.action = na.pass)
par(mfrow = c(1,1))
```

## Ljung-Box test 
From the plot it is shown that we have autocorrelation of residuals however there is a better tool to check autocorrelation which is 
Ljung-Box test for max 10 lags

```{r}
Box.test(resid(arima1),
         type = 'Ljung-Box',
         lag = 10)
```


H0 for Ljung-Box is that, Residuals are fine, means there is not auto correlation!
we used lag = 10 means, there is no autocorrelation - order = 1, order = 2.....order = 10 as our p value is 0.9932 we fail to reject H0 which means our model is valid!however to obtan more percise model i can restrict my model further by droping statisticly insiginificant models and then we can compare which one is better model

```{r}
arima1_1 <- Arima(covid$deaths,
                    order = c(5,1,5),
                    fixed = c(0,0,0,0,NA,  # vector of the same
                              NA,0,0,0,0,NA), # length as the total number of parameters
                    include.constant = TRUE  # last is for the intercept (if included)
)   

summary(arima1_1)
coeftest(arima1_1)


Box.test(resid(arima1_1),
         type = 'Ljung-Box',
         lag = 10)
```

## Model Copmarision 

Now as we have 2 models and 2 of them seem valid then lets compare them for this we will do AIC and BIC

### AIC

```{r}
AIC(arima1, arima1_1)  
```

### BIC

```{r}
BIC(arima1, arima1_1)
```

Furthermore there is also a way to create the best model for this we can use auto.arima() function lets try and then we can compare all the models 


```{r}
auto_arima.AIC <- auto.arima(covid$deaths,
                             d = 1,             # parameter d of ARIMA model
                             max.p = 10,        # Maximum value of p
                             max.q = 10,        # Maximum value of q
                             max.order = 14,    # maximum p+q
                             start.p = 1,       # Starting value of p in stepwise procedure
                             start.q = 1,       # Starting value of q in stepwise procedure
                             ic = "aic",        # Information criterion to be used in model selection.
                             stepwise = FALSE,  # if FALSE considers all models
                             allowdrift = TRUE, # include a constant
                             trace = TRUE)      # show summary of all models considered

```

### The result might be surprising

```{r}
coeftest(auto_arima.AIC)
```


Okay we have arima1 and arima1_1 models so we can compare them for comparing the model we use AIC abd BIC

AIC better than for the best manually selected model

```{r}
AIC(auto_arima.AIC)
```

BIC worse than for the best manually selected model
```{r}
BIC(auto_arima.AIC)
```


### Ljung-Box test
```{r}
Box.test(resid(auto_arima.AIC),
         type = "Ljung-Box", lag = 10)
```
After testing our model time to see how our model fits with the real data

```{r}

regressionMetrics <- function(real, predicted) {
    MSE <- mean((real - predicted)^2) # Mean Squera Error
    RMSE <- sqrt(MSE) # Root Mean Square Error
    MAE <- mean(abs(real - predicted)) # Mean Absolute Error
    MAPE <- mean(abs(real - predicted)/abs(real)) # Mean Absolute Percentage Error
    MedAE <- median(abs(real - predicted)) # Median Absolute Error
    TSS <- sum((real - mean(real))^2) # Total Sum of Squares
    RSS <- sum((predicted - real)^2)  # Explained Sum of Squares
    R2 <- 1 - RSS/TSS
    
    result <- data.frame(MSE, RMSE, MAE, MAPE, MedAE, R2)
    return(result)
}
```


## Lets plot that 

```{r}
fit.r <- covid$deaths  # real values - last 31 observations, last observation iis zero thats why i dint take it
fit.m <- xts(fitted(arima1),order.by=index(covid))# forecast
colors_ <- c("black", "red")
plot.xts(merge(fit.r, fit.m), plot.type="single",
         col = colors_,   # colors for subsequent lines
         ylab = "Daily number of deaths", # axes labels
         main = "Real data Vs Model data") # title above the plot
addLegend("topleft", c("real","model"), text.col = colors_)
```
## Preparing the fprecast 

### Time to start ppreparing for the forecastusing our model


lets see last 25% observations

```{r}
tail(covid, 104)


covid.sample <- covid["/2021-01-07", 1]

tail(covid.sample)
```

## FORECAST for death
### model arima515 
Estimating the model on shorter sample


```{r}
arima515s <- Arima(covid.sample$deaths,  # variable
                   order = c(5,1,5),   # (p,d,q) parameters
                   include.constant = T)

arima515s
```

## Prediction

### Forecasting using the model

```{r}
forecast515  =  forecast(arima515s, # model for prediction
                         h = 104) # how many periods outside the sample
```

## Results
```{r}
forecast515
```

The forecasts are indexed with a observation number,not a date!

```{r}

class(forecast515$mean)
# it is a ts object, not xts


# it includes:
# Point Forecast - predicted values
forecast515$mean

# 80% and 95% confidence intervals
forecast515$lower
forecast515$upper
```


If we want to easily put together both real data and the forecast on the plot, we have to convert both to ts or both to xts objects


```{r}

class(covid)

tail(covid, 104) # indexed by date

# ts() function does easily convert xts to ts object

class(ts(covid))

tail(ts(covid), 104) # indexed by observation number
```

## Plot with the Forecast 

### Long term forecast 


```{r}
# original data
plot(ts(covid[,1]), main = "104 day forecast of death",xlim = c(0,418), ylim = c(0, 3000))
# line at the end of a sample period (start of the forecast)
# observation number 388
abline(v = 313, lty = 2, col = "gray")
# line for the forecast
lines(forecast515$mean, col = "red", lwd = 2)
# line for the 95% confidence interval
# indexed by observation numbers from the forecast
lines(313:416,forecast515$lower[,2], col = "red", lty = 3)
lines(313:416,forecast515$upper[,2], col = "red", lty = 3)
```

As we can the out forecast is very poor and not suffiecent. However we can try forecasting with our model for shorter time period
Now will try with shorter time period means also with fewer observaions 



```{r}

tail(covid, 31)

# we cut last 31 observations (since 2021-03-22 )
# and keep them aside as the out-of-sample
# to perform a forecasting exercise
# and have a possibility to assess it

covid.sample <- covid["/2021-03-22", 1]

tail(covid.sample)

```

## FORECAST for death 

### Estimating  the model on shorter sample

```{r}
arima515s <- Arima(covid.sample$deaths,  # variable
                   order = c(5,1,5),   # (p,d,q) parameters
                   include.constant = T)

arima515s
```

## Prediction short term
```{r}
forecast515  =  forecast(arima515s, # model for prediction
                         h = 31) # how many periods outside the sample
```

## Result short term

```{r}
forecast515
```

The forecasts are indexed with a observation number, not a date!

```{r}
class(forecast515$mean)
# it is a ts object, not xts


# it includes:
# Point Forecast - predicted values
forecast515$mean

# 80% and 95% confidence intervals
forecast515$lower
forecast515$upper
```


If we want to easily put together both real data and the forecast on the plot, we have to convert both to ts or both to xts objects

```{r}
class(covid)

tail(covid, 31) # indexed by date
```

ts() function does easily convert xts to ts object

```{r}
class(ts(covid))

tail(ts(covid), 31)
```

## Plot with the Forecast 

```{r}
plot(ts(covid$deaths), main = "31 day forecast of death")
# line at the end of a sample period (start of the forecast)
# observation number 388
abline(v = 388, lty = 2, col = "gray")
# line for the forecast
lines(forecast515$mean, col = "red", lwd = 2)
# line for the 95% confidence interval
# indexed by observation numbers from the forecast
lines(388:418,forecast515$lower[,2], col = "red", lty = 3)
lines(388:418,forecast515$upper[,2], col = "red", lty = 3)
```

### Zoomed plot

```{r}
plot(ts(covid[,1]), main = "31 day forecast of death",
     xlim = c(388,418), ylim = c(0, 500))
abline(v = 388, lty = 2, col = "gray")
lines(forecast515$mean, col = "red", lwd = 2)
lines(388:418,forecast515$lower[,2], col = "red", lty = 3)
lines(388:418,forecast515$upper[,2], col = "red", lty = 3)
```

## Quality Check

Checking forecast quality:for simplicity of the following formulas
lets define two new objects:

```{r}

death.r <- tail(covid$death[-417], 31)  # real values - last 31 observations, last observation iis zero thats why i dint take it
death.f <- as.numeric(forecast515$mean) # forecast

death_forecast <- data.frame(death.r, death.f)
```

### Basis of different measures 
lets add the basis for different measures of the forecast error

```{r}
death_forecast$mae  =  abs(death.r - death.f)
death_forecast$mse  =  (death.r - death.f)^2
death_forecast$mape  =  abs((death.r - death.f)/death.r)
death_forecast$amape  =  abs((death.r - death.f)/(death.r+death.f))
```

### Average Error 

```{r}
colMeans(death_forecast[,3:6])
```

## Conclusion

Our forecast on short time period seems to be more accurate. However as our forecast is not as accurate we as supposed, we cannot verify our main hypothesis that death cases will decline significantly 
NOTE: Despite performing better in the sample, does not mean the Model will necessarily be the best for forecasting


## Bibliography 

Although, we used different literature as a source to compare our work and to check our step by step approach, everything is done according to our lectures and labs. To avoid complication and further misunderstanding we didnt tend to use other ways or methodologies rather than what we have been taught in the classes. However, to check and compare our results many sources have been used, which I mentioned in the literature review! 