---
title: "Classification"
author: "Elbrus Gasimov"
date: "5/31/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## introduction


Our dataset contains transactions made by credit cards in September 2013 by European cardholders. 
This dataset presents transactions that occurred in two days, where we have 492 frauds out of 284,807 transactions. The dataset is highly unbalanced, the positive class (frauds) account for 0.172% of all transactions.

Features V1, V2, … V28 are the principal components obtained with PCA, the only features which have not been transformed with PCA are 'Time' and 'Amount'. Feature 'Time' contains the seconds elapsed between each transaction and the first transaction in the dataset. The feature 'Amount' is the transaction Amount, this feature can be used for example-dependant cost-sensitive learning. Feature 'Class' is the response variable and it takes value 1 in case of fraud and 0 otherwise.

Given the class imbalance ratio, we recommend measuring the accuracy using the Area Under the Precision-Recall Curve (AUPRC). Confusion matrix accuracy is not meaningful for unbalanced classification.


### Importing the required packages and libraries


```{r}
#install.packages(c('ROCR','ggplot2','corrplot','caTools','class',
#                   'randomForest','pROC','imbalance'))
library(ROCR)
library(ggplot2)
library(corrplot)
library(caTools)
library(class)
library(randomForest)
library(pROC)
library(imbalance)
library(rpart)
```

# The data is saved as a csv file namely creditcard.csv

## Importing the data

```{r pressure, echo=FALSE}
transaction <- read.csv('creditcard.csv')
```


## About the Dataset:
 The dataset includes 284807 observations and 31 variables.
 Out of these 31 variables, one variable (Class) is the variable of interest.
 The dataset contains transactions made by credit cards in September 2013 by 
 European cardholders over a span of two days.
 Due to confidentiality issues, dataset contains only numerical input variables
 which are the result of a PCA transformation. Only features which have not
 been transformed with PCA are 'Time' and 'Amount'. Feature 'Time' contains the
 seconds elapsed between each transaction and the first transaction in the
 dataset. The feature 'Amount' is the transaction Amount.

## Basic Exploratory Data Analysis

Viewing some initial observations of the dataset


```{r}
head(transaction)
```

## Viewing Structure of transactions

```{r}

str(transaction)

sapply(transaction, FUN=class)
```


## After the basic data exploration we found that all of the variables including the target variable are of numeric data type, due to the PCA.

## Viewing the Frequency distribution of "Class" Variable

```{r}
table(transaction$Class)
prop.table(table(transaction$Class))
ggplot(data = transaction,aes(x=Class))+geom_bar(col="slateblue3") 
```

## The dataset is highly imbalanced, the positive class (frauds) account for mere 0.172% of all transactions. This is really good for the credit card
company but not for this analysis.

## Plotting the Amount - Fraud chart

```{r}
ggplot(data = transaction,aes(y=Amount,x=Class))+geom_point(col="slateblue3") +
  facet_grid(~Class)
```

## This plot tells us that all the fraudulent transactions were made for less
## than around 2500, which is considerably less than that of true transactions.

## Exploring Time Column with Regards to the Class variable

```{r}
ggplot(data = transaction,aes(x=Time),fill=factor(Class))+
  geom_histogram(col='red',bins=48) +  facet_grid(Class ~ ., scales = 'free_y')

```


## As the unit of measure for Time here is seconds, it makes it harder to 
## understand and get anything useful out of this.

# Converting Time(Seconds) into Hours for better understanding

```{r}
transaction$hours <- round(transaction$Time/3600)
unique(transaction$hours)
```

## Checking correlation between the variables

```{r}
transaction$Class <- as.numeric(transaction$Class)
corr <- cor(transaction[],method="pearson")
corrplot(corr,method = "circle", type = "lower")
```


## This shows that there is not much correlation among the variables.
## "V2" has some negative correlation with "Amount". However, we can ignore it.

## Converting "Class" to factor type for Modeling purposes

```{r}
transaction$Class <- as.factor(transaction$Class)
```


## Splitting the dataset into the Training dataset and Test dataset
## We are doing 75-25 split..


```{r}
# For reproducability of the random selection
set.seed(123) 
```

## We are using sample.split and subset functions from caTools Library to split
## the data randomly


```{r}
split = sample.split(transaction$Class, SplitRatio = 0.75)
training_set = subset(transaction, split == TRUE)
test_set = subset(transaction, split == FALSE)

```

## Verify if the proportion of Class is still the same

```{r}
table(training_set$Class)
prop.table(table(training_set$Class))

table(test_set$Class)
prop.table(table(test_set$Class))
```

## To balance out the minority(fraud) and majority class in dataset, we are using
## the SMOTE technique. We are setting ratio at 0.8 which means we are
## oversampling our minority class to come upto 80% of the majority class. 
## It will maintain the ratio and the machine learning model will now be able to
## easily understand the minority class also.

```{r}
smoted_training_set<-oversample(dataset= training_set, 
                                method = 'SMOTE',ratio = 0.8)
table(smoted_training_set$Class) 
prop.table(table(smoted_training_set$Class)) 

colnames(smoted_training_set)

```

## Excluding the Class variable to calculate the scaled values for each columns

```{r}
smoted_training_set[-31] <- scale(smoted_training_set[-31])
test_set[-31] <- scale(test_set[-31])

```


## Fitting Logistic Regression to the Training set

```{r}

classifier <- glm(formula = Class ~ .,
                  family = binomial,
                  data = smoted_training_set)


summary(classifier)

```

## Summary of this model clearly tells that almost all of the variables are 
## statistically significant to the model. Only V24 is there which is not 
## significant. We can ignore this and keep it with us as the value is slightly
## higher than the 5% level of significance.

## Predicting the Test set results

```{r}
prob_pred <- predict(classifier, type = 'response', newdata = test_set[-31])
```

## Setting a condition that even if there are more than 40% chances(>0.4) of 
## getting a Fraud Transaction, model should predict it as Fraud.

```{r}
y_pred <- ifelse(prob_pred > 0.4, 1, 0)
```

## Calculating and Plotting the AUC-ROC curve
## AUC(Area under the curve) curve tells that how accurately our model is
## predicting together with its overall Performance.
## The thumb-rule is that Higher the AUC Better our model is at predicting.

```{r}
calc_auc <- function (prob_pred,test_y){
  predict_log <- prediction(prob_pred,test_y) # Prediction Probability
  
  #Calculating ROC Curve...
  table(test_y, prob_pred>0.4)
  roc_curve<- performance(predict_log,"tpr","fpr")  
  # plot(roc_curve)
  plot(roc_curve, colorize=T)
  
  #Calculating AUC and printing....
  auc<- performance(predict_log,"auc")
  paste(auc@y.values[[1]])
  
}

```


# Make confusion matrix

```{r}
cm <- table(test_set[, 31], y_pred > 0.4)
cm
```


## Print AUC

```{r}
print(calc_auc(prob_pred,test_set$Class))

```


## The value of AUC is 0.97 which is really high. However, we cannot just decide
## if it is the best model for our dataset or not.

## Creating Random Forest Model
## ntree = 100

```{r}
set.seed(123)
forest.classifier = randomForest(x = smoted_training_set[-31],
                                 y = smoted_training_set$Class,
                                 ntree = 100)
```

## Predicting Classes using the model


```{r}
forest.y_pred = predict(forest.classifier, newdata = test_set[-31])
```

## Predicting Probabilities

```{r}
forest.y_prob = predict(forest.classifier, newdata = test_set[-31], type='prob')
```

## Predicting Log for AUC Generation

```{r}
forest.predict_log <- prediction(as.numeric(y_pred),as.numeric(test_set$Class)) 
```

## Calculating AUC Curve

```{r}
forest.auc<- performance(forest.predict_log,"auc")
print( (forest.auc@y.values[[1]]))
```

## The AUC value in this case is 0.52, which means the logistic regression is far
## better than Random forest model for this dataset.

## Making the confusion matrix

```{r}
cm <- table(test_set[, 31], forest.y_pred)
cm
```

## Creating AUC ROC Curve

```{r}
roc_curve<- performance(forest.predict_log,"tpr","fpr")  
plot(roc_curve, colorize=T)
```


## Decision Tree Classifier

```{r}
tree.model <- rpart(Class~.,smoted_training_set,method='class')
tree.predict <- predict(tree.model, newdata = test_set[-31],type = 'class')
tree.predict_log <- prediction(as.numeric(tree.predict),
                               as.numeric(test_set$Class))
```


## Calculating AUC Curve

```{r}
tree.auc<- performance(tree.predict_log,"auc")
print( (tree.auc@y.values[[1]]))
```


## The AUC value comes out to be 0.68 which is significantly lower than that of 
## logistic regression.

# Conclusions:

Logistic Regression works well for this dataset.
We are having 0.97 of AUC for Logistic Regression which is a significant 
performance of the model.
There can be seen a rise in the Fraud transactions during the morning or earlymorning time period.
Frauds were not of high values in comparison to the normal transactions.
Maximum amount of fraud transaction was 2500-3000 in the native currency.
