---
title: "Movie Recommendation Dashboard"
runtime: shiny
output: 
  html_document
---
# Using Shiny And R Markdown.
### *TOOL#4, TOOL#5*.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### R Markdown Shiny scripting.
This R Markdown script is used to generate Dashboard reports for Movie Recommendataion system.
&nbsp; 

&nbsp;

&nbsp;


```{r echo=FALSE, message = FALSE, warning = FALSE}
## R libraries.
library(utils)
library(reshape2)
library(recommenderlab)

library(ggplot2)                       
library(data.table)
library(reshape2)
```

## Movie Ratings Dashboard Reports.
&nbsp; 

&nbsp;

&nbsp;


```{r echo=FALSE, message = FALSE, warning = FALSE}
rep_rating_data <-read.csv("ratings.csv")

ratingMatrix <- dcast(rep_rating_data, userId~movieId, value.var = "rating", na.rm=FALSE)
ratingMatrix <- as.matrix(ratingMatrix[,-1]) #remove userIds
#Convert rating matrix into a recommenderlab sparse matrix
ratingMatrix <- as(ratingMatrix, "realRatingMatrix")
```

### Similarity Matrix of User Ratings
```{r echo=FALSE, message = FALSE, warning = FALSE}
similarity_mat <- similarity(ratingMatrix[1:4, ],
                               method = "cosine",
                               which = "users")

image(as.matrix(similarity_mat), main = "User's Rating Similarities")
```
&nbsp; 

&nbsp;

&nbsp;


### Heat Map of Movie Ratings of first 25 users.
```{r echo=FALSE, message = FALSE, warning = FALSE}
image(ratingMatrix[1:20, 1:25], axes = FALSE, main = "Heatmap of first 25 x 25 Movie Ratings")

movie_ratings <- ratingMatrix[rowCounts(ratingMatrix) > 50,
                             colCounts(ratingMatrix) > 50]
```
&nbsp; 

&nbsp;

&nbsp;


### Heat Map of Movie Ratings for Top Users.
```{r echo=FALSE, message = FALSE, warning = FALSE}
minimum_movies<- quantile(rowCounts(movie_ratings), 0.98)
minimum_users <- quantile(colCounts(movie_ratings), 0.98)
image(movie_ratings[rowCounts(movie_ratings) > minimum_movies,
                     colCounts(movie_ratings) > minimum_users],
                     main = "Heatmap of Top Users and Movies Ratings")
```
&nbsp; 

&nbsp;

&nbsp;


### Map of Normalised Ratings.
```{r echo=FALSE, message = FALSE, warning = FALSE}
normalized_ratings <- normalize(movie_ratings)
#sum(rowMeans(normalized_ratings) > 0.00001)

image(normalized_ratings[rowCounts(normalized_ratings) > minimum_movies,
                          colCounts(normalized_ratings) > minimum_users],
main = "Normalized Ratings of the Top Users")
```


